Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-05 paper/paper_draft.qmd
Updates to introduction and abstract

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral
    cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming
    connectivity was found to influence reef temporal behaviour at 15m depth. Our findings suggest that
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the
    performance of deep reefs, contrasting current intervention
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have
    important implications for reef management, indicating that depth refugia should be
    prioritized for further consideration in conservation planning alongside connectivity
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Coral reef ecosystem resilience - the capacity to maintain coral populations and avoid
ecosystem regime shifts - is conceptualised to depend on three interconnected mechanisms.
First, the ability of individual corals to survive thermal stress. Second, the connectivity
networks that enable recolonisation after disturbances [@hock2017; @mumby2008; @torres2024].
Third, the maintenance, and accretion, of the carbonate framework that provides reef
structure [@perry2012]. We aim to assess how these mechanisms interact under projected
climate conditions to identify the characteristics of resilient reefs through scenario-based
coral population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of
thousands of individual reefs interconnected through larval dispersal patterns mediated by
hydrodynamic, spatial, and ecological processes [@cappo2000]. The GBR faces severe projected
coral cover declines under climate change [@bozec2025; @bozec2022; @condie2021]. Coral cover
declines will be compounded by ocean acidification, tropical cyclones, and crown-of-thorns
starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011;
@wolff2018]. The 2023-24 Summer mass bleaching event may be signalling a shift towards this
reef future, with some level of bleaching recorded on 79% of surveyed reefs (total 1001
surveyed reefs) [@australianinstituteofmarinescience2024; @cantin2024].

At the individual coral scale, survival during thermal stress events depends on both
environmental and biological factors. Environmental factors can either amplify or reduce
thermal stress exposure. Depth is a crucial factor that can provide a refuge from thermal
stress [@álvarez-noriega2018; @baird2018; @smith2014]. Biological factors including coral
size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and thermal tolerance
through acclimation and inheritance [@humanes2022; @lachs2024] all influence survival
outcomes. Importantly, corals demonstrate adaptive capacity to develop thermal tolerance
that could delay widespread bleaching under moderate emission scenarios [@bairos-novak2021;
@humanes2022], representing a crucial mechanism for reef persistence.

Beyond individual survival, reef-scale resilience depends on connectivity networks that
enable recolonization after disturbances [@angeler2023; @elmhirst2009; @hock2017]. However,
connectivity creates both strengths and vulnerabilities. Larval source populations that
export more larvae than they import are essential for system-wide recovery, providing larvae
that enable recolonization at distant locations [@hock2017; @thomas2015]. Yet these source
reefs are paradoxically more vulnerable individually, relying heavily on local recruitment
for their own recovery. When disturbances impact major source populations, cascading effects
reduce larval export and slow recovery throughout the connected network [@figueiredo2022;
@hock2017; @munday2009].

Long-term reef viability requires not only coral survival and recolonization, but also
maintenance of the physical reef structure through positive carbonate budgets. Scleractinian
corals build the primary habitat framework, but reef persistence depends on the balance
between carbonate accumulation [@perry2015] and loss through bioerosion, physical damage
[@kenyon2023], and dissolution [@eyre2014; @eyre2018]. This balance is influenced by coral
cover and community complexity [@perry2012; @desbiens2021], environmental processes, and
ocean chemistry [@perry2015; @eyre2018].

Recent advances in whole-ecosystem larval connectivity modelling **\[Insert GBR1
connectivity citation.\]** and coral thermal response research [@bairos-novak2021] enable
integrated investigation of how multiple factors interact to determine climate resilience.
By incorporating the latest coral population models [@ribeirodealmeida2024], ecosystem-wide
connectivity data, and updated climate projections [@dixon2022], we can examine which
factors most strongly influence reef persistence under different scenarios. While
connectivity has been emphasized as critical for GBR resilience [@hock2017], and remains an
important consideration, our integrated modelling reveals depth plays a fundamental role in
determining maintenance of reef coral populations and positive carbonate budgets under
current projections of climate change.

>>>>

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral
    cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    reefs were analysed for reef coral cover time series and their ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming
    connectivity was found to influence reef temporal behaviour at 15m depth. These findings suggest that
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the
    performance of deep reefs, contrasting current intervention
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have
    important implications for reef management, indicating that depth refugia should be
    prioritized for further consideration in conservation planning alongside connectivity
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Coral reef ecosystem resilience - the capacity to maintain coral populations and avoid
ecosystem regime shifts - is conceptualised to depend on three interconnected mechanisms.
First, the ability of individual corals to survive thermal stress. Second, the connectivity
networks that enable recolonisation after disturbances [@hock2017; @mumby2008; @torres2024].
Third, the maintenance, and accretion, of the carbonate framework that provides reef
structure [@perry2012]. This investigation aims to assess how these mechanisms interact
under projected climate conditions to identify the characteristics of resilient reefs in the
Great Barrier Reef (GBR) ecosystem.

The GBR is the world's largest coral reef ecosystem consisting of thousands of individual
reefs interconnected through larval dispersal patterns mediated by hydrodynamic, spatial,
and ecological processes [@cappo2000]. The GBR faces severe projected coral cover declines
under climate change [@bozec2025; @bozec2022; @condie2021]. Coral cover declines will be
compounded by ocean acidification, tropical cyclones, and crown-of-thorns starfish outbreaks
[@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24
Summer mass bleaching event may be signalling a shift towards this reef future, with some
level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs)
[@australianinstituteofmarinescience2024; @cantin2024].

At the individual coral scale, survival during thermal stress events depends on both
environmental and biological factors that can either amplify or reduce thermal stress. Depth
is a crucial factor that can provide a refuge from thermal stress [@álvarez-noriega2018;
@baird2018; @smith2014]. Biological factors including coral size [@brandt2009], taxa
[@álvarez-noriega2018; @mcclanahan2005], and thermal tolerance through acclimation and
inheritance [@humanes2022; @lachs2024] all influence survival outcomes. Corals demonstrate
adaptive capacity to develop thermal tolerance that could delay widespread bleaching under
moderate emission scenarios [@bairos-novak2021; @humanes2022], representing a crucial
mechanism for reef persistence.

Beyond individual survival, reef-scale resilience depends on connectivity networks that
enable recolonization after disturbances [@angeler2023; @elmhirst2009; @hock2017]. Larval
source populations that export more larvae than they import are essential for system-wide
recovery, providing larvae that enable recolonization at distant locations [@hock2017;
@thomas2015]. Yet these source reefs are paradoxically more vulnerable individually, relying
heavily on local recruitment for their own recovery. When disturbances impact major source
populations, cascading effects reduce larval export and slow recovery throughout the
connected network [@figueiredo2022; @hock2017; @munday2009].

Long-term reef viability requires not only coral survival and recolonization, but also
maintenance of physical reef structures through positive carbonate budgets. Scleractinian
corals build the primary habitat framework, but reef persistence depends on the balance
between carbonate accumulation [@perry2015] and loss through bioerosion, physical damage
[@kenyon2023], and dissolution [@eyre2014; @eyre2018]. This balance is influenced by coral
cover and community complexity [@perry2012; @desbiens2021], environmental processes, and
ocean chemistry [@perry2015; @eyre2018].

Recent advances in whole-ecosystem larval connectivity modelling [@choukroun2025] and coral
thermal response research [@bairos-novak2021] enable integrated investigation of how
multiple factors interact to determine climate resilience. By incorporating the latest coral
population models [@ribeirodealmeida2024], ecosystem-wide connectivity data, and updated
climate projections [@dixon2022], we can examine which factors most strongly influence reef
persistence under different scenarios. While connectivity has been emphasized as critical
for GBR resilience [@hock2017], and remains an important consideration, integrated modelling
reveals depth plays a fundamental role for maintenance of reef coral populations and
positive carbonate budgets under current projections of climate change.
